---
title: "test_gsusie"
author: "Yixin (Izzy) Zhang"
date: "2024-03-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Read ATAC data
```{r}
library(vroom)
library(dplyr)

Sys.setenv("VROOM_CONNECTION_SIZE"=1000000000)

atac_act_file_path <- "./GSE162170/GSE162170_multiome_atac_gene_activities.tsv.gz"

# Read data with vroom, skipping the first row (since we don't have data at [1,1] it woudl crunch the cols)
atac_real_data <- vroom(atac_act_file_path, delim = "\t", col_names = FALSE, col_types = cols(), skip = 1)
```
transpose data, and set genenames as colnames
```{r}
atac_real_data = data.frame(t(atac_real_data))
# atac_real_data = setNames(data.frame(t(atac_real_data)), atac_real_data[,1])
```

```{r}
colnames(atac_real_data) <- unlist(atac_real_data[1, ])
```

```{r}
atac_real_data <- atac_real_data[-1, ]
```

```{r}
# atac_real_data <- mutate_all(atac_real_data, function(x) as.numeric(as.character(x)))
atac_real_data[] <- lapply(atac_real_data, function(x) if(is.character(x)) as.numeric(x) else x)


```

```{r}
percentage_zeros <- sapply(atac_real_data, function(column) {
  mean(column == 0) * 100
})
```

```{r}
hist(percentage_zeros)
```

34 non-unique col names, make them unique
```{r}
library(dplyr)

# Create a function to rename duplicates with a unique suffix
make_unique <- function(names) {
  counts <- ave(seq_along(names), names, FUN = seq_along)
  unique_names <- ifelse(counts == 1, names, paste(names, counts, sep = "_"))
  return(unique_names)
}

# Apply the function to your dataframe's column names
colnames(atac_real_data) <- make_unique(colnames(atac_real_data))

```

Remove cols with missingness over 50% (arbitrary) (80% too much memory during gsusie)
```{r}
library(dplyr)

# Calculate the percentage of 0s for each column
percent_zeros <- atac_real_data %>%
  summarise(across(everything(), ~mean(.x == 0) * 100))

# Identify columns with 50% or fewer zeros
vars_to_keep <- names(percent_zeros)[percent_zeros <= 50]

# Filter the original dataframe to keep only the identified columns
atac_real_data_filtered <- atac_real_data %>%
  select(all_of(vars_to_keep))

# View the filtered dataframe
# print(atac_real_data)

```

```{r}
cols_to_remove <- sapply(atac_real_data, function(column) mean(column == 0) > 0.5)
atac_real_data_0_filtered <- atac_real_data[, !cols_to_remove]
```

prepare RNAseq data
get from ``test_data``
```{r}
rna_7_var <- read.csv('./test_data/rna_7_var.csv')
```

take a look at distribution
```{r}
hist(rna_7_var$SOX2, breaks = 12)
```
apply log1p
```{r}
SOX2RNA <- exp(log1p(rna_7_var$SOX2))
hist(SOX2RNA)
```
```{r}
# write.csv(atac_real_data, "./test_data/rearranged_atac_activity.csv")
write.csv(atac_real_data_0_filtered, "./test_data/zero_rm_50_atac_activity.csv")
```

```{r}
mean(SOX2RNA)
var(SOX2RNA)
```
overdispersion

# Apply gsusie
```{r}
atac_real_data_0_filtered <- read.csv("./test_data/zero_rm_50_atac_activity.csv")
```

```{r}
atac_matrix <- as.matrix(atac_real_data_0_filtered)
colnames(atac_matrix) <- NULL
rna_matrix <- as.matrix(SOX2RNA)
```

Took 15 minutes to run on 50% of ATACseq data
```{r}
library(gsusie)
res_gs <- gsusie(cbind(atac_matrix, 1), rna_matrix, family = "poisson")
```

```{r}
print_gsusie_coefficients(res_gs)
```


```{r}
print(res_gs$sets)
```

```{r}
gsusie_plot(res_gs, y = "PIP", include_intercept = FALSE, intercept_index = (ncol(atac_matrix)+1))
```

```{r}
gsusie_plot(res_gs, y = "PIP", include_intercept = TRUE)
```

find out the names to those atacseq
```{r}
atac_filtered_colnames <- colnames(atac_real_data_0_filtered)
atac_filtered_colnames[as.integer(res_gs$sets$cs)]
```

## ATACseq data contains missing (0s) and is lognormal 
1. gsusie does not work with -Inf values (Too much abnormal subjects detected!)
2. replace 0 with NA (Too much abnormal subjects detected!)
```{r}
atac_matrix_NA <- atac_matrix
atac_matrix_NA[atac_matrix_NA == 0] <- NA
atac_log_matrix_NA <- log(atac_matrix_NA)
```
standardize
```{r}
atac_log_matrix_NA_st <- atac_log_matrix_NA %>% mutate_all(~(scale(.) %>% as.vector))
```

3. listwise deletion & pairwise deletion (removes all the rows)
```{r}
# rows_with_zeros <- apply(atac_matrix, 1, function(x) any(x == 0))
# atac_matrix_0_rm <- atac_matrix[!rows_with_zeros, ]
# atac_log_matrix_0_rm <- log(atac_matrix_0_rm)
```
4. Imputation
```{r}
library(hdImpute)
## hdImpute: fast random forest
hdImpute(data.frame(atac_log_matrix_NA_st), batch = 100, seed = 0, save = TRUE)
```

```{r}
res_gs <- gsusie(cbind(atac_log_matrix_0_rm, 1), rna_matrix, family = "poisson")
```

